---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.cap = "",
  dpi = 96
)
```

# plantstressR

<!-- badges: start -->
[![R-CMD-check](https://github.com/agrobioestat/plantstressR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/agrobioestat/plantstressR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`plantstressR` answers one question:

> **What is the physiological stress signature of a plant, and how do its traits
> interact?**

You bring one tidy table of physiological traits measured under a control
treatment and one or more stress treatments. The package does the rest: it
puts every trait on a common signed scale, integrates them into a single
rankable index, draws the signature, and maps the network of traits that
respond together.

## Installation

```r
# install.packages("remotes")
remotes::install_github("agrobioestat/plantstressR")
```

## The workflow

| Function | What it gives you |
|---|---|
| `validate_stress_data()` | Design checks before anything is computed |
| `calculate_sri()` | A signed, standardized Stress Response Index per trait |
| `integrated_stress_index()` | One weighted index per genotype, plus a ranking |
| `plot_stress_signature()` | The signature as a heat map or a radar plot |
| `stress_network()` | A regularized partial-correlation network with trait modules |
| `stress_module_scores()` | The modules scored with the trait indices |
| `stress_tolerance_index()` | The classical selection indices (`STI`, `SSI`, `GMP`, `TOL`, ...) |
| `stress_ordination()` | The sample-level view of the trait space |
| `stress_index_ci()` | Bootstrap limits and rank stability for the ranking |
| `run_plantstress_app()` | The whole workflow in a Shiny dashboard, no code |

Every design function takes `block =` for randomized complete block designs.

## Point and click

If you would rather not write the script at all:

```r
# install.packages("shiny")
run_plantstress_app()
```

The dashboard loads your CSV, asks which columns are the treatment, the
genotype, the block and the traits, and then gives you the same signature,
ranking, network, ordination and tolerance indices the functions below
produce — every panel is computed by the exported functions themselves, so
anything you see can be reproduced from a script.

## Example

```{r example}
library(plantstressR)

traits <- c("Fv_Fm", "PIabs", "DI0_RC", "A", "gs", "Ci", "RWC", "SPAD", "shoot_biomass")

sri <- calculate_sri(
  brachiaria_stress,
  treatment = "drought_level",
  control = "control",
  traits = traits,
  by = "genotype",
  block = "block",
  verbose = FALSE
)

summary(sri)
```

Positive values always mean *stress impaired this trait*, whichever direction
the trait moves. `Fv/Fm` falling and `DI0/RC` rising are both damage, and
`trait_directions()` keeps that bookkeeping straight.

`block = "block"` is what makes this a blocked analysis: each trait is fitted
as `trait ~ treatment + block` within the genotype, so the variation between
replicates leaves the scale instead of diluting the effect. Drop the argument
and you get the completely randomized reading of the same table.

Rank the genotypes:

```{r isi}
integrated_stress_index(sri, weights = "precision")
```

But is that order real? Resample the trial and find out:

```{r ci}
stress_index_ci(sri, n_boot = 200, seed = 1)
```

`p_best` is the share of resamples in which a genotype came out the most
tolerant one. A ranking whose winner sits at `p_best = 0.35` is a ranking you
should not select on.

Draw the signature:

```{r signature, fig.height = 5}
plot_stress_signature(sri, units = "G1")
```

Map how the traits interact:

```{r network}
net <- stress_network(
  brachiaria_stress,
  traits = traits,
  treatment = "drought_level",
  level = c("moderate", "severe"),
  block = "block",
  sri = sri
)

net
```

```{r network-plot, fig.height = 5}
plot(net, color_by = "sri")
```

And score those modules with the signature:

```{r module-scores}
stress_module_scores(net, sri)
```

Cross-check against the classical selection indices:

```{r sti}
stress_tolerance_index(
  brachiaria_stress,
  trait = "shoot_biomass",
  treatment = "drought_level",
  control = "control",
  by = "genotype",
  indices = c("STI", "GMP", "SSI")
)
```

## Learn more

```r
vignette("plantstressR")
```

## Citation

```r
citation("plantstressR")
```

## License

GPL (>= 3)
