---
title: "Multivariate phenotyping of plant stress with plantstressR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multivariate phenotyping of plant stress with plantstressR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

## The question this package answers

A stress phenotyping trial produces a wide table: dozens of physiological
traits, measured on the same plants, under a control treatment and one or more
stress treatments. The analytical question is almost always the same one:

> **What is the physiological stress signature of this material, and how do the
> traits interact?**

`plantstressR` answers exactly that question and nothing else. It does not
classify, predict or integrate omics layers. It takes one tidy table and returns
four things: a signed index per trait, one integrated index per genotype, a
picture of the signature, and the network of traits that respond together.

```{r load}
library(plantstressR)

data(brachiaria_stress)
str(brachiaria_stress[, 1:8])
```

Three trials ship with the package, all simulated and documented as such:

| Data set | What it is for |
|---|---|
| `brachiaria_stress` | Four genotypes, three drought levels, four blocks, 29 traits |
| `wheat_salinity` | Six cultivars at three sites, for the multi-environment functions |
| `maize_heat` | Small and deliberately messy, to see what the design checks say |

The input contract is deliberately small:

| Column kind | Requirement |
|---|---|
| Treatment | One column, one level flagged as the control |
| Grouping (optional) | One column, typically genotype; every level needs its own control |
| Traits | Numeric columns, any units, any number |

Everything else in the table is ignored.

## Step 1 — validate the design before computing anything

`validate_stress_data()` is the gate. It fails loudly on problems that make the
indices meaningless (an unknown control level, a genotype without controls, a
fully missing trait) and warns about problems that only make them fragile (low
replication, constant traits, heavy missingness).

```{r validate}
traits <- c(
  "Fv_Fm", "PIabs", "DI0_RC", "A", "gs", "Ci",
  "RWC", "SPAD", "shoot_biomass"
)

validate_stress_data(
  brachiaria_stress,
  treatment = "drought_level",
  control = "control",
  traits = traits,
  by = "genotype"
)
```

## Step 2 — one signed index per trait

`calculate_sri()` converts each trait into a **Stress Response Index**: the
standardized difference between a stress treatment and its own control,
re-signed so that a positive value always means *the stress impaired this
trait*.

Sign handling is the part that is easy to get wrong by hand. `Fv/Fm` falling and
`DI0/RC` rising both mean damage; averaging them raw would cancel real
information. `trait_directions()` resolves the orientation, using a small
dictionary of well-known physiological traits and falling back to
"higher is better".

```{r directions}
trait_directions(traits)
```

```{r sri}
sri <- calculate_sri(
  brachiaria_stress,
  treatment = "drought_level",
  control = "control",
  traits = traits,
  by = "genotype",
  method = "glass"
)

summary(sri)
```

Choosing `method`:

| `method` | Scale | Use it when |
|---|---|---|
| `"glass"` | SD of the control | Default. The stress treatment inflates variance and you do not want that to shrink the index |
| `"cohen"` | Pooled SD | Both groups have comparable dispersion |
| `"hedges"` | Pooled SD, bias corrected | Few replicates per plot |
| `"relative"` | Control mean | You want a percentage change, on ratio-scale traits only |

### Trials laid out in blocks

Most field and glasshouse trials are not completely randomized: plants sit in
benches, strips or runs, and a whole replicate can be slightly more vigorous
than the others for reasons that have nothing to do with the treatment. That
between-block spread lands in the standard deviation that scales the index,
and a diluted scale means a diluted signature.

Naming the block column is all it takes. Each trait is then fitted as
`trait ~ treatment + block` within the unit, the difference between adjusted
means becomes the numerator, and the residual — block-free — standard
deviation becomes the scale:

```{r sri-block}
sri_rcbd <- calculate_sri(
  brachiaria_stress,
  treatment = "drought_level",
  control = "control",
  traits = traits,
  by = "genotype",
  block = "block"
)

summary(sri_rcbd)
```

The treatment difference is a property of the data and does not move. What
changes is the scale it is divided by, so the same experiment speaks louder
once the blocks are accounted for:

```{r block-comparison}
data.frame(
  design = c("completely randomized", "randomized complete block"),
  mean_abs_sri = c(
    mean(abs(sri$sri), na.rm = TRUE),
    mean(abs(sri_rcbd$sri), na.rm = TRUE)
  )
)
```

`stress_network()` and `stress_ordination()` take the same argument, and
`validate_stress_data()` will tell you when the layout cannot support it — for
instance when a treatment sits inside a single block, where its effect and the
block's effect are the same number.

## Step 3 — see the signature

```{r heatmap, fig.height = 5}
plot_stress_signature(sri, units = "G1")
```

Rows are ordered by hierarchical clustering of their index profiles, so traits
that respond alike sit next to each other. Stars mark adjusted p-values.

For a small, curated panel in a manuscript figure, the radar view reads better:

```{r radar, fig.height = 5}
plot_stress_signature(sri, type = "radar", units = "G1", top_n = 6)
```

## Step 4 — one number per genotype, and a ranking

`integrated_stress_index()` collapses the per-trait indices into a single
weighted score per genotype and stress level. Because the trait indices are
already dimensionless and consistently signed, the weighted mean is a
meaningful severity score.

```{r isi}
isi <- integrated_stress_index(sri, weights = "precision")
isi
```

Ranks and the `0-100` rescaling are computed **within** each stress level:
moderate and severe deficit are different environments, and mixing them in one
ranking would compare genotypes across environments they did not share.

The weights and the per-trait contributions are recoverable, which is what makes
a ranking defensible in a paper:

```{r weights}
stress_weights(isi)

head(stress_contributions(isi), 5)
```

Choosing `weights`:

| `weights` | Meaning |
|---|---|
| `"equal"` | Every trait counts the same; use with an a priori trait panel |
| `"precision"` | Traits measured on fewer plants are down-weighted |
| `"pca"` | Traits carrying the dominant axis of stress variation weigh more |
| named vector | Expert weights, rescaled to sum to one |

A word on `"precision"`. Standardizing a trait has already divided out how
noisy the instrument is, so with balanced replication and similar dispersion in
both groups `"precision"` lands on the same answer as `"equal"`. Two things
make it differ, and both are worth down-weighting: a trait measured on fewer
plants than the others — the usual fate of gas exchange next to SPAD — and a
trait whose variance the stress treatment blew up, which is measured much less
precisely relative to its own scale even when every plant was read.

Weighting by the *full* standard error of the index instead would be a mistake:
that quantity grows with the effect size itself, so it would quietly demote the
very traits that responded to the stress. `se_sampling` exists precisely to
separate the two.

`plot()` gives the two figures that belong together: the ranking, and the
decomposition that justifies it.

```{r isi-plot, fig.height = 3.6}
plot(isi)
```

```{r isi-contribution, fig.height = 3.6}
plot(isi, type = "contribution")
```

## Step 4b — is the ranking real?

A table of genotypes in order looks like a decision, but it is an estimate.
Run the trial again and the middle of the table would shuffle. Before selecting
on a ranking, ask how much of it survives resampling:

```{r ci}
# 100 resamples keeps this vignette quick; use 500 or more for real work.
ci <- stress_index_ci(sri_rcbd, n_boot = 100, seed = 1)
ci
```

Three things to read here. `conf_low` and `conf_high` bracket the index itself.
`rank_low` and `rank_high` bracket the positions the genotype occupied across
the resamples — when they span half the table, that genotype's position is not
information. And `p_best` is the share of resamples in which it came out first:
the direct answer to "is this really my most tolerant material?".

```{r ci-plot, fig.height = 3.6}
plot(ci)
```

One caution about what is being resampled. A standardized index is a difference
divided by a standard deviation, and that denominator is an estimate too. Draw
a handful of control plants and it can collapse — in this trial the smallest
control standard deviation, `Fv/Fm` at `0.0027`, falls to `0.0006` in the worst
resample. Five times a smaller ruler means five times a bigger index, and the
upper limit of the interval runs away. So `scale = "fixed"`, the default, holds
each trait at the scaling the observed trial gave it and lets only the response
vary. Use `scale = "resampled"` if you want the stricter reading, and expect
long upper tails.

## Step 4c — more than one site, or more than one year

Everything so far describes a single trial. A breeding program rarely stops
there: the same genotypes are grown at several sites, or repeated across
seasons, and the question becomes whether a ranking obtained in one place
travels to the next.

Name both columns in `by`. Each genotype is then indexed at each site against
the control of that site, which is the right comparison — a drought treatment
in a dry year is not the same treatment as in a wet one:

The package ships such a trial: `wheat_salinity`, six cultivars grown at three
sites under sodium chloride, in complete blocks within each site.

```{r met}
data(wheat_salinity)

met_sri <- calculate_sri(
  wheat_salinity,
  treatment = "salinity",
  control = "control",
  traits = c("Fv_Fm", "A", "RWC", "Na", "K", "MDA", "grain_yield"),
  by = c("cultivar", "site"),
  block = "block",
  verbose = FALSE
)

stability <- stress_stability(integrated_stress_index(met_sri))
stability
```

Note that nothing had to be said about `Na` or `K`. Sodium accumulating is
damage and potassium falling is damage, and `trait_directions()` resolves both
from its dictionary — which is the bookkeeping this package exists to keep
straight.

Three columns carry the message. `sd_isi` is how much a genotype's index moved
between sites. `rank_min` and `rank_max` are the best and worst positions it
occupied — when they span most of the table, that genotype's ranking is not
portable. And `ecovalence_pct` is Wricke's share of the genotype-by-environment
interaction, so the material responsible for the instability of the whole trial
can be named rather than guessed at.

```{r met-plot, fig.height = 3.6}
plot(stability)
```

Bottom left is where you want to be: little affected, and the same everywhere.

## Step 5 — how the traits interact

A signature says *what* changed. The network says *what changed together*.
`stress_network()` estimates partial correlations, so each edge is the
association between two traits after removing every other trait in the panel;
the dense webs of indirect correlation that make raw correlation networks
unreadable simply disappear.

```{r network}
net <- stress_network(
  brachiaria_stress,
  traits = traits,
  treatment = "drought_level",
  level = c("moderate", "severe"),
  sri = sri
)

net
```

```{r network-plot, fig.height = 5}
plot(net, color_by = "sri")
```

Modules become interpretable once they are scored with the trait indices.
`stress_module_scores()` turns a nine-trait signature into three sentences:

```{r module-scores}
stress_module_scores(net, sri)
```

Two modeling choices matter here:

* **Which plants.** A network estimated on stressed plants only describes the
  stressed physiology. Estimated on the whole trial, it mostly recovers the
  treatment contrast.
* **Regularization.** The correlation matrix is shrunk towards the identity
  before inversion. `lambda = "auto"` picks the smallest intensity that yields a
  well-conditioned matrix, which keeps the estimate usable when the number of
  traits approaches the number of plots.

## Step 6 — the sample-level view

`calculate_sri()` summarizes treatments and `stress_network()` summarizes
traits. When the individual plants themselves are the object of interest,
`stress_ordination()` places them in the trait space.

```{r ordination, fig.height = 5}
ord <- stress_ordination(
  brachiaria_stress,
  treatment = "drought_level",
  traits = traits
)

ord
plot(ord)
```

## Step 7 — the agronomic cross-check

A physiological ranking is more convincing when it agrees with the classical
selection indices, and more interesting when it does not.
`stress_tolerance_index()` computes them from a single productivity trait.

```{r sti}
sti <- stress_tolerance_index(
  brachiaria_stress,
  trait = "shoot_biomass",
  treatment = "drought_level",
  control = "control",
  by = "genotype",
  indices = c("STI", "GMP", "SSI", "TOL", "YSI")
)

sti
```

No single index is right on its own. `TOL` and `SSI` reward stability even when
it comes with low potential; `MP` and `GMP` reward potential even when the loss
is large. `rank_mean` averages the ranks after orienting every index towards
tolerance, and `rank_overall` ranks the genotypes on that average.

Comparing the two rankings is the point:

```{r cross-check}
merge(
  isi[isi$group == "severe", c("unit", "isi", "rank")],
  sti[sti$group == "severe", c("unit", "STI", "rank_overall")],
  by = "unit"
)
```

## The whole workflow

```{r workflow, eval = FALSE}
sri <- calculate_sri(data,
  treatment = "trt", control = "ctrl",
  by = "genotype", block = "block"
)
isi <- integrated_stress_index(sri, weights = "equal")
net <- stress_network(data,
  treatment = "trt", level = "stress",
  block = "block", sri = sri
)

plot(sri)                          # the signature
plot(isi)                          # the ranking
plot(net)                          # the interactions
stress_module_scores(net, sri)     # the modules, scored
```

One table in, four answers out.

## Without writing any of it

The same workflow is available as a Shiny dashboard, which is often the
quickest way to explore a new trial or to hand the analysis to someone who
does not work in R:

```{r app, eval = FALSE}
# install.packages("shiny")
run_plantstress_app()
```

Load a CSV, point the app at the treatment, genotype, block and trait columns,
and every tab is computed by the exported functions used above — so any result
seen on screen can be reproduced by a script.

```{r session}
sessionInfo()
```
